Efficient pedigree recording for fast population genetics simulation

### 2018-06-06 review-responses.tex

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We'd like to thank the reviewers for a careful reading of the paper;
as detailed below, their helpful comments have resulted in a substantial amount of clarification
about the overall structure and purpose of the paper.
We hope you agree that our changes have addressed their comments and concerns.

@@

\begin{quote}
    Both reviewers are clearly very enthusiastic about this manuscript, but both
    have made suggestions that require major revisions to address. Please be
    very careful in addressing the following points exhaustively:
    \begin{itemize}
        \item  Discuss the limitations imposed by Wright-Fisher assumptions (Rev 1) 
            and how the proposed approach may/may not be extended 
            to other demographic and population scenarios.

        \item  Improve the readability of the text and provide necessary background information (Rev 1)

        \item  Discuss implementation availability (Rev 2)
    \end{itemize}
\end{quote}

We thank the reviewers for helpful feedback, and respond in detail to their points below.
In part to address all of these points,
we have tried to clarify the main purpose of this paper, which is
to describe algorithms and data structures for tree sequences,
and how these are useful in simulation.
Our main goal is not to write a standalone simulation package.
What we do is useful and important, because: 
(a) we provide a fully functional and flexible, well-documented API
so that developers can include this idea in their own simulator
(with no restrictions beyond conserved synteny of chromosomes); and
(b) we explain the data structures,
which are closely tied to the idea of a tree sequence,
and therefore important for even end users of a simulator to understand.
We describe and implement Wright--Fisher simulations not because of any limitation of our method,
but to keep things as simple as possible, only focusing on those aspects necessary 
for recording the tree sequence.
We'd also like to note that we are developing a set of tutorials,
at \url{https://github.com/tskit-dev/tutorials}.

@@

We'd like to emphasize that although the tools for dealing with tree sequences
are currently bundled with the coalescent simulator, \msprime{},
this is for historical reasons only, 
and we are in the process of separating out these tools into a separate package,
to be called \tskit.
The format for storing tree sequences was developed for the coalescent simulator,
but the work we describe here generalized the data structure and algorithms
to allow it to be used in any simulator.
We have stated this more clearly in the text at \llname{msprime_rel}.

TODO: think about switching to \tskit{} in the paper?

\begin{point}{}
    Overall, I find the method very clever and appealing, and I am convinced that
    it will be well received by the community. However, nowadays, researchers
    generate and use genomic data in organisms other than humans and from wild
    populations that by far are not Wright-Fisher populations. How are you selling
    your method to that broader category of researchers? Compared to SPLATCHE,
    which simulates structured populations with gene flow, it may look like a
    regression (quantiNemo implemented something similar, but I don't think it was
    ever published).
    \emph{[\ldots]}
    As a developer, I am interested in the limitations and the constraints that may
    prevent me to implement your method to model populations that are not W-F.
    populations. In particular, you didn't address the question of structured
    populations with gene flow, or the effect of overlapping generations. Although
    developing an approach that is efficient in structured populations is probably
    out of scope here, giving a minimum of guidelines or just discussing the issue
    would greatly improve the scope of the manuscript.
\end{point}

\reply{
    We're sorry to have not made this more clear,
    but our goal was \emph{not} to make a faster WF simulator,
    but rather to describe a scheme that can be easily applied to \emph{any} simulator
    whatsoever, as long as it has the notion of a recombining chromosome.
    We have added a note to this effect before the proof-of-concept WF algorithm \llname{ll:not_wf};
    and substantially expanded the discussion of what steps need to be taken in general \revref.
    Also see the summarizing paragraph at \llname{ll:other_words}.
    The only thing that should stand in your way as a developer
    to implementing this is bookkeeping;
    and we've provided tools to make this bookkeeping easier
    (like the \texttt{sort tables} functionality,
    so that the simulator does not have to guarantee sortedness of the tables itself).
}

@@

\begin{point}{}
    More generally, I found the ms very technical and addressed to a specialized
    audience. I would not assume, as you do, that readers are necessarily
    proficient in coalescent theory. Therefore, please improve the introduction by
    providing more information on: \emph{(points below)}
\end{point}

\reply{
    Thanks for the suggestion;
    since this paper is entirely about forwards simulation,
    we hope not to depend on coalescent theory at all in any crucial way.
    We have added some more explanation of the relationship to simplification
    at \revref;
    and have tried to better signpost the technical arguments at \llname{ll:coal_signpost}.
}

\begin{point}{}
    The idea of running simulations forward in time for the demography and then
    backward in time for the neutral genetics is by far not new and was already
    implemented in the software SPLATCHE (Ray, Currat, Foll, Excoffier,
    Bioinformatics 2010, 26: 2993-2994), although probably not as efficiently. The
    reference should be added.
\end{point}

\reply{
    Thanks for drawing our attention to the (good) analogy 
    between the forward-backward scheme used by SPLATCHE
    and the process of tree sequence simplification,
    which we had not considered.
    While there are indeed deep connections between this approach and tree sequence simplification, 
    they are however fundamentally different:
    the coalescent is a stochastic process that is dual to the forwards-time process,
    and relies on certain assumptions for correctness;
    whereas tree sequence simplification is deterministic (just ``bookkeeping'').
    We have updated the introduction to discuss this connection with SPLATCHE \revref,
    and have discussed further the relationship of coalescence to simplification \revreffull{1}{2}.
}

@@

\begin{point}{}
  What is the use of gene trees? What can be inferred/calculated from them that
  we cannot get from the endpoint genetic polymorphism data? (An example comes
very late in the ms, should be moved up.)
\end{point}

\reply{
    We agree that discussion of uses of gene trees would make nice motivation,
    but think it is somewhat outside of the scope of this paper:
    for our purposes here, it suffices to explain that they help everything go faster.
    We've expanded the discussion \revref.
}

@@

    In addition, the use of the compressed format of the sequence tree is quite
    amazing. The 144k over the VCF equivalent is unbelievable.
\end{quote}

Thanks!

\begin{point}{}
    I only see one major criticism: accessibility. The authors do not provide a
    clear overview of how to obtain/install/use the algorithms presented. A
    ``proof-of-concept implementation'' is really not sufficient. In contrast,
    msprime has great documentation, but the current website does not seem to
    describe the forward simulation model (in contrast to what is suggested by the
    manuscript). \url{https://github.com/ashander/ftprime} does have installation
    instructions, but seems to be entirely geared toward integration with simuPOP.
    Most of the figures are generated with fwdpp, which are supposedly included
    here (\url{https://github.com/petrelharp/ftprime_ms}), but it is not at all clear how
    one would access the code given the current organization of this repository. It
    is unrealistic to expect all downstream users to code up the algorithms
    themselves (and would result in code that is not as well tested). The authors
    should have a more dedicated repository for users to access the software and
    start running their own simulations.
\end{point}

\reply{
    As the reviewer suggests, our main goal here is to provide easy access
    to the \emph{algorithm}, not to a forward simulation model --
    we aren't describing a forward simulation model, 
    but rather a way to speed up and get more information out of an existing one.
    But, is the algorithm accessible?
    An interface to our algorithms is available at the python level or at the C level.
    The python interface is documented through the \msprime{} documentation
    (but you need to select ``latest'': \url{https://msprime.readthedocs.io/en/latest/});
    and closely mirrors the underlying C library.
    We have also provided a link to the  developing set of tutorials 
    at \url{https://github.com/tskit-dev/tutorials} \llname{ll:tutorials}.
}

@@

\begin{point}{}
The Appendices seem to be out of order with the text. The current order does
not seem any more logical.
\end{point}

\reply{
    Good point. Fixed.
}

\begin{point}{}
In Figure 1, it would potentially be helpful to see y-axis on a log scale so
that we could see scaling on the low end. Legend for this figure refers to
``right column of panels''. How many threads were used to obtain the dashed line?
\end{point}

\reply{
    Thanks for the suggestions: we've switched the figure to a log scale,
    and noted the number of threads (four).
    The speedup is not that impressive for a variety of reasons, though, which
    are hard to tease apart, so we haven't elaborated on this point.
    (But, for the reviewer's interest: false sharing is one possible
    issue, but perhaps $5 \times 10^4$ diploids may be too few to make
    parallelization ``worth'' it when spawning new threads each generation.)
}

\begin{point}{}
``its{'}'' on page 11.
\end{point}

### 2018-06-06 forwards_paper.tex
update

\begin{abstract}
    In this paper we describe how to efficiently record the entire genetic history of a population
    in forwards-time, individual-based population genetics simulations with arbitrary breeding models,
    population structure and demography.
    This approach dramatically reduces the computational burden of tracking individual genomes
    by allowing us to simulate only those loci that may affect reproduction
    (those having non-neutral variants).
    The genetic history of the population is recorded as a succinct tree sequence
    as introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    Recording the results of each breeding event
    results in a linear growth in the amount of information that must be stored.
    However, there is a great deal of redundancy in this information,
    since the majority of individuals simulated over the course of the simulation
    will leave no descendents in the eventual sample.
    We solve this storage problem by providing an algorithm to quickly `simplify'
    a tree sequence by removing this irrelevant history for a given set of genomes.
    By periodically simplifying the history with respect to the extant population,
    we show that the total storage space required is modest
    and overall large efficiency gains can be made over classical forward-time simulations.
    We implement a general-purpose framework for recording and simplifying genealogical data,
    which can be used to make simulations of any population model more efficient.
    We modify two popular forwards-time simulation frameworks to use this new approach
    and observe efficiency gains in large, whole-genome simulations of one to two orders of magnitude.
    In addition to speed, our method for recording pedigrees has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
        can be stored in $O(N log N + M)$ space,
        making it feasible to store a simulation's entire final generation
        (and its history) rather than only a subset.
    (3) A simulation can easily be initialized with a more efficient coalescent simulation 
        of deep history.
\end{abstract}

@@

In this paper, we describe a storage method for \emph{succinct tree sequences}
(and hence, genome sequence) as well as an algorithm for simplifying these.
The data structure is \emph{succinct} in the sense that its space usage is close to optimal,
while still allowing efficient retrieval of information (see, e.g., \citet{gog2014theory}).
We also describe how these tools can efficiently record,
and later process, the embellished population pedigree from a forwards-time simulation.
While providing substantial savings in computational time and space, our methods provide 
in principle much more information than simply simulating the genomes -- the tree sequence
encodes all marginal genealogies of individuals living at the end of the simulation.
These marginal genealogies enable fast data storage and processing,
but also provide additional information that can be used to better understand
the notoriously complex dynamics of population genetics. \revpoint{1}{4}
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.
This work originated as improvements to the algorithmic tools and data structures 
in the coalescent simulator \msprime{}; \linelabel{msprime_rel}
the tools are currently available bundled with the Python package \msprime{},
but will soon be separately available as a Python package named \tskit{}.

@@

\begin{figure}
    \includegraphics[]{sims/rawspeed_logy}
    \caption{
        \label{fig:runtimes_selection}
    Total run time per single simulation replicate as a function of region length.
    % measured as the scaled recombination parameter $\rho = 4Nr$.
    Line color represents different diploid population sizes ($N$).
    The left figure shows run times for standard simulations including neutral mutations.
    The right column shows run times of simulations that recorded the pedigree
    and added neutral mutations afterwards.
    The dashed line in the right panel shows results for an implementation using \fwdpy{}
    where the pedigree simplification steps were handled in a separate thread of execution
    and fitness calculations were parallelized.
    Simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
    }
\end{figure}

@@

The \msprime\ Python Tables API provides a general purpose toolkit for importing
and processing succinct tree sequences. Interoperation with Python simulators
is then straightforward.  The implementation we benchmark here uses
\texttt{pybind11} (\url{https://github.com/pybind/pybind11/}) to interface
with the \fwdpp{} \cpp{} API \citep{fwdpp}. No modifications were
required to the \fwdpp{} code base; rather, we simply need to bookkeep parent/offspring labels,
and perform simple processing of the recombination breakpoints from each mating
event to generate node and edge data. This information is then periodically copied
to the \msprime\ Tables API, where it is sorted and simplified.

>>>>

\begin{abstract}
    In this paper we describe how to efficiently record the entire genetic history of a population
    in forwards-time, individual-based population genetics simulations with arbitrary breeding models,
    population structure and demography.
    This approach dramatically reduces the computational burden of tracking individual genomes
    by allowing us to simulate only those loci that may affect reproduction
    (those having non-neutral variants).
    The genetic history of the population is recorded as a succinct tree sequence
    as introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    Recording the results of each breeding event
    results in a linear growth in the amount of information that must be stored.
    However, there is a great deal of redundancy in this information,
    since the majority of individuals simulated over the course of the simulation
    will leave no descendents in the eventual sample.
    We solve this storage problem by providing an algorithm to quickly `simplify'
    a tree sequence by removing this irrelevant history for a given set of genomes.
    By periodically simplifying the history with respect to the extant population,
    we show that the total storage space required is modest
    and overall large efficiency gains can be made over classical forward-time simulations.
    We implement a general-purpose framework for recording and simplifying genealogical data,
    which can be used to make simulations of any population model more efficient.
    We modify two popular forwards-time simulation frameworks to use this new approach
    and observe efficiency gains in large, whole-genome simulations of one to two orders of magnitude.
    In addition to speed, our method for recording pedigrees has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
        can be stored in $O(N \log N + M)$ space,
        making it feasible to store a simulation's entire final generation
        (and its history) rather than only a subset.
    (3) A simulation can easily be initialized with a more efficient coalescent simulation 
        of deep history.
\end{abstract}

@@

In this paper, we describe a storage method for \emph{succinct tree sequences}
(and hence, genome sequence) as well as an algorithm for simplifying these.
The data structure is \emph{succinct} in the sense that its space usage is close to optimal,
while still allowing efficient retrieval of information (see, e.g., \citet{gog2014theory}).
We also describe how these tools can efficiently record,
and later process, the embellished population pedigree from a forwards-time simulation.
While providing substantial savings in computational time and space, our methods provide 
in principle much more information than simply simulating the genomes -- the tree sequence
encodes all marginal genealogies of individuals living at the end of the simulation.
These marginal genealogies enable fast data storage and processing,
but also provide additional information that can be used to better understand
the notoriously complex dynamics of population genetics. \revpoint{1}{4}
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.
This work originated as improvements to the algorithmic tools and data structures 
in the coalescent simulator \msprime{}; \linelabel{msprime_rel}
the tools are currently available bundled with the Python package \msprime{},
but will soon be separately available as a package named \tskit{}, 
which will include a Python API and an embeddable C library.

@@

\begin{figure}
    \includegraphics[]{sims/rawspeed_logy}
    \caption{
        \label{fig:runtimes_selection}
    Total run time per single simulation replicate as a function of region length.
    % measured as the scaled recombination parameter $\rho = 4Nr$.
    Line color represents different diploid population sizes ($N$).
    The left figure shows run times for standard simulations including neutral mutations.
    The right column shows run times of simulations that recorded the pedigree
    and added neutral mutations afterwards.
    The dashed line in the right panel shows results for an implementation using \fwdpy{}
    where the pedigree simplification steps were handled in a separate thread of execution
    and fitness calculations were parallelized across four cores.
    Simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
    }
\end{figure}

@@

The \msprime{} Python Tables API provides a general purpose toolkit for importing
and processing succinct tree sequences, and a collection of tutorials are being developed
at \url{https://github.com/tskit-dev/tutorials}. \linelabel{ll:tutorials}
Interoperation with Python simulators is then straightforward.  
The implementation we benchmark here uses \texttt{pybind11} 
(\url{https://github.com/pybind/pybind11/}) to interface with the \fwdpp{} \cpp{} API \citep{fwdpp}. 
No modifications were
required to the \fwdpp{} code base; rather, we simply need to bookkeep parent/offspring labels,
and perform simple processing of the recombination breakpoints from each mating
event to generate node and edge data. This information is then periodically copied
to the \msprime\ Tables API, where it is sorted and simplified.
